Double Standards in Judging Collective Action\thanks{}

### 2024-11-05 Documents/manuscript.tex
2nd revision

\noindent Collective action is a powerful force driving social change
but often sparks contention about what actions are acceptable means to
effect social change. Five studies (total \(N = 2,979\)) investigated
double standards in judging collective action---that is, whether
observers judge the same protest actions to be more acceptable depending
on who the protesters are and what they are protesting. Two studies used
item response theory to develop an instrument of 25 controversial
protest actions to measure where people draw the line between acceptable
and unacceptable forms of collective action. Three preregistered
experiments found no consistent evidence for ingroup bias in terms of
social class when judging protests for workers' rights (Experiment 1),
in terms of race when judging protests for and against defunding the
police (Experiment 2), and in terms of gender when judging protests for
and against restricting abortion (Experiment 3). Instead, we found that
participants who identified as left-wing/liberal (Experiments 1--3) and
rejected system-justifying beliefs (Experiments 1--2) considered the
same protest actions more acceptable when a cause aligned with their
ideological orientation (for workers' rights, for defunding the police,
against restricting abortion) than when it did not (against defunding
the police, for restricting abortion). Conservative participants
considered the same actions somewhat more acceptable when protesters
supported, rather than opposed, restricting abortion (Experiment 3) but
considered all protest actions, for and against defunding the police,
equally unacceptable (Experiment 2). Our findings have theoretical and
practical implications for understanding the often-divided response to
social movements.

@@

\begin{center}
\noindent 11,654 words
\end{center}

@@

\noindent This research shows that what people consider an acceptable
\emph{means} of protest depends on what the \emph{end} of a protest is.
That is, observers sometimes judge the \emph{same} controversial protest
action (e.g., blocking a road) to be more acceptable when the
protesters' cause (e.g., defunding the police) aligns with their own
ideological position. Conservatives, who support social order and oppose
social change, tend to consider all protest actions less acceptable than
progressives. Progressives, in turn, tend to consider disruptive but
non-violent actions acceptable means to achieve social change and
challenge social injustice. This finding helps explain the often-divided
response to recent protest movements.

@@

Another explanation for ideology-based double standards is that they are
rooted in ideological differences in how conservatives and progressives
relate to social change and social inequality which, in turn, are
motivated by epistemic, existential, and relational needs (Jost et al.,
2003; Jost, 2021). As motivated social cognition, ideological
differences shape how we think and feel about political issues, actions,
and events. In this view, ideology-based double standards depend on
whether the cause of a protest aligns with someone's ideological
positions.

@@

Item response theory is a conceptual and statistical framework for
understanding how the characteristics of both items and respondents
shape responses to a set of items (DeMars, 2010). When respondents judge
whether each of a set of actions is an acceptable means of protest (1 =
\emph{yes}, 0 = \emph{no}), a two-parameter logistic item response
theory model estimates responses as a function of several latent
(unobserved) parameters:
\[ \Pr ( y_\textit{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) \]
where \(\Pr ( y_\textit{ij} = 1 )\) is the probability that respondent
\(j\) considers action \(i\) an acceptable means of protest,
\(\theta_j\) estimates how \emph{accepting} respondent \(j\) is of
various protest actions, \(\beta_i\) estimates how \emph{acceptable}
action \(i\) is, and \(\alpha_i\) estimates how \emph{discriminating}
action \(i\) is.

Figure \ref{fig:f1} illustrates those relationships for three
hypothetical protest actions. \(\theta_{(z)}\) is the
\emph{z}-standardized tendency for a participant to be more accepting of
various actions so that \(\beta_i\) determines the probability that the
average respondent (\(\theta_{(z)} = 0\)) considers an action an
acceptable means of protest. Action 2 (\(\beta_2 = 1.28\)) is more
acceptable than Actions 1 and 2 (\(\{ \beta_2, \beta_3 \} = 0\)) as the
average respondent is more likely to consider the former (\(\Pr = .78\))
than the latter (\(\Pr = .50\)) acceptable means of protest. Action 3
(\(\alpha_3 = 3.60\)) is more discriminating than Actions 1 and 2
(\(\{ \alpha_1, \alpha_2 \} = 1\)) as the slope for the relationship
between how accepting a respondent is and how likely they are to
consider an action acceptable is steeper for the former action. This
means that knowing whether someone considers Action 3 acceptable
provides more information about them than knowing whether they consider
Actions 1 or 2 acceptable.

\begin{figure}
\centering
\caption{Item response curves for three hypothetical protest actions}
\includegraphics[scale=1]{../Scale Development/figures/figure-1.png}
\caption*{\textit{Note.} $\theta_{(z)}$ follows a standard normal distribution and is shown as both $z$-scores (bottom) and percentiles (top).}
\label{fig:f1}
\end{figure}

In this research, item response theory serves two purposes. First, it
provides a statistical framework for scale development. In Studies 1 and
2, we select the most discriminating and informative actions to build an
instrument for capturing double standards in judging collective action.
Second, it provides a formal and statistical definition of double
standards in judging collective action as respondents being more
accepting of the same protest actions depending on who the protesters
are and what they are protesting. In Experiments 1--3, we operationalize
double standards as differences in \(\theta_{(z)}\) which, as
\(\theta_{(z)}\) is \(z\)-standardized, correspond to Cohen's \(d\)
effect sizes.

@@

\noindent Participants were then asked to name between five and ten
actions that fit that definition. Participants were encouraged to think
of actions that varied in how acceptable they were in their opinion. We
recoded responses into a smaller set of unique actions, which we
supplemented with protest actions from the psychological and political
science literature (e.g., Sharp, 1973). This process resulted in 72
actions that varied in how acceptable we would expect them to be. For
details, see Supplemental Online Material (SOM).

@@

We estimated a graded response model (Bürkner, 2021; Samejima,
1997)---an item response theory model for ordinal response
variables---for participants' ratings of how often an action would be an
acceptable form of collective action.\footnote{Ratings of how often a
  given action would be an acceptable form of collective action were
  strongly and negatively correlated with how disruptive, violent, and
  extreme participants considered the action to be and how negative they
  felt about it (Table S3).} Based on the results, we selected the 25
most discriminating, informative, and relevant protest actions to form
an instrument for measuring double standards in judging collective
action in Experiments 1--3. We selected controversial, and thus
diagnostic, actions (e.g., disrupting traffic) over actions that almost
all respondents considered acceptable (e.g., signing petitions) or
unacceptable (e.g., political violence; Figure \ref{fig:f2}). For
detailed ratings and results, see SOM.

@@

By varying the participants' and the protesters' group memberships, we
tested the preregistered hypothesis that participants would judge the
same protest actions as more acceptable if the protesters were ingroup
members protesting for the ingroup's rights than if they were outgroup
members protesting for the outgroup's rights. By comparing reactions to
working-class and middle-class protesters, we further tested how
observers judge collective action by lower-status (i.e.~working-class)
and higher-status (middle-class) group members.\footnote{For Experiments
  1 and 2, we had preregistered the additional hypotheses that all
  participants would judge the same actions as more acceptable when the
  protesters were from the higher-status group rather than the
  lower-status group (Experiment 1) and when the goal of the protest was
  to support rather than challenge the system (Experiment 2). These
  hypotheses reflected the assumption that the motivation to justify the
  system was, to some extent, universal. With hindsight, we no longer
  think that this assumption is tenable and, therefore, omit the
  hypotheses based on it from the revised manuscript. We still report
  the same preregistered analyses and their results.} In
non-preregistered analyses, we tested whether participants would judge
the same protest actions as more or less acceptable depending on their
own ideological positions in terms of political orientation,
system-justifying beliefs, and social dominance orientation.\footnote{In
  Experiment 1, we explored a broader range of operationalizations of
  political ideology, including social dominance orientation, than in
  the other experiments.} In this way, Experiment 1 provided a first
test of the hypothesized class- and ideology-based double standards in
judging collective action.

@@

We preregistered the sample size as well as all hypotheses,
inclusion/exclusion criteria, measures, and manipulations
(\url{https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b}).
We made all materials, data, and analysis scripts available online
(\url{https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62}).
We followed sample size recommendations for item response theory models
(DeMars, 2010, p. 36), planning to recruit 500 participants.

@@

In the screening survey, participants read an accessible definition of
what working-class and middle-class/professional jobs are (for details,
see SOM). Participants then answered, among other questions, whether
they considered their current job---or the jobs they had had in the past
or expected to have in the future---to be a working-class job or a
middle-class/professional job. As preregistered, we excluded
participants from the lower-status group who did not respond
``working-class job'' and participants from the higher-status group who
did not respond ``middle-class/professional job''.

@@

\begin{figure*}[!t]
\centering
\caption{Results from the preregistered (\textbf{A}) and non-preregistered (\textbf{B}) analyses for Experiment 1}
\includegraphics[scale=1]{../Experiment 1/figures/figure-3}
\caption*{\textit{Note.} HS = Higher Status, LS = Lower Status. $\theta_{(z)}$ is the \textit{z}-standardized tendency to consider more controversial actions acceptable means of protest. (\textbf{A}) $p$ and $n$ are, respectively, the predicted proportion and number of actions a participant would consider acceptable means of protest in each condition. Bars enclose the 95\% most plausible estimates. (\textbf{B}) Ribbons enclose, from the darkest to the lightest shade, the 50\%, 80\%, 95\%, and 99\% most plausible estimates.}
\label{fig:f3}
\end{figure*}

@@

We estimated this model in CmdStan (Gabry \& Cesnovar, 2021; Stan
Development Team, 2021) using Bayesian statistical methods. Bayesian
inference involves choosing a likelihood function and prior
distributions. A likelihood function links the observed data to the
model parameters and states how likely the observed data are given
different values of said model parameters. Prior distributions state how
plausible different values of said model parameters are before
considering the observed data. Bayesian inference applies Bayes' theorem
to update prior distributions in light of the observed data to produce
posterior distributions. In contrast to \(p\)-values and confidence
intervals, posterior distributions have a straight-forward
interpretation as stating how plausible different values of the model
parameters are given the observed data.

Our model derived the likelihood of participants' responses from a
Bernoulli likelihood function with a logistic regression equation
linking the two item parameters (\(\alpha_i\), \(\beta_i\)), the one
participant parameter (\(\theta_j\)), and the three regression
coefficients to the observed data. To identify the model, we constrained
\(\theta_j\) to have a mean of zero and constrained \(\alpha_i\) to have
a fixed mean and to be non-negative. We used partial pooling to estimate
\(\alpha_i\), \(\beta_i\), and \(\theta_j\). Our model assigned weakly
informative prior distributions (Gelman et al., 2017) to all model
parameters.\footnote{Our model assigned \(\text{Half-Cauchy} (0, 3)\)
  prior distributions to the standard deviations of \(\alpha_i\),
  \(\beta_i\), and \(\theta_j\) and \(\text{Normal} (0, 3)\) prior
  distributions to all other model parameters.} We report point
estimates, based on the median of posterior samples, and uncertainty
intervals, based on the quantiles of posterior samples, that enclose the
95\% most plausible estimates.

@@

Figure \ref{fig:f3} shows (B) the \(z\)-standardized propensity to
consider more controversial protest actions acceptable as a function of
the three ideological orientation variables. We found that participants
who reported a more right-wing political orientation
(\(\beta_{xy} = -0.25, [-0.37, -0.15]\)) and who expressed more
agreement with system-justifying beliefs
(\(\beta_{xy} = -0.20, [-0.30, -0.10]\)) tended to find fewer collective
actions to be acceptable. In contrast, we found that, after controlling
for the other two variables, social dominance orientation was not
associated with participants' judgments about how acceptable various
collective actions are (\(\beta_{xy} = -0.05, [-0.15, 0.05]\)). Overall,
these findings suggest that people who are right-wing and endorse
system-justifying beliefs tend to find various collective actions to be
less acceptable than people who are left-wing and do not endorse
system-justifying beliefs.\footnote{We ran additional analyses to
  explore how these associations differed across experimental
  conditions. We found uncertainty intervals for these associations to
  overlap across conditions, suggesting that we did not have enough data
  to differentiate these varying effects.}

@@

\begin{figure*}[!t]
\centering
\caption{Results from the preregistered analyses for Experiment 2}
\includegraphics[scale=1]{../Experiment 2/figures/figure-4}
\caption*{\textit{Note.} Against = Protesters oppose defunding the police. For = Protesters support defunding the police. $\theta_{(z)}$ is the \textit{z}-standardized tendency to consider more controversial actions acceptable means of protest. Ribbons enclose, from the darkest to the lightest shade, the 50\%, 80\%, 95\%, and 99\% most plausible estimates.}
\label{fig:f4}
\end{figure*}

\begin{figure*}[!t]
\centering
\caption{Predictions from the preregistered (system justification) and non-preregistered (political orientation) analyses for Experiment 2}
\includegraphics[scale=1]{../Experiment 2/figures/figure-5}
\caption*{\textit{Note.} Against = Protesters oppose defunding the police. For = Protesters support defunding the police. $\theta_{(z)}$ is the \textit{z}-standardized tendency to consider more controversial actions acceptable means of protest.}
\label{fig:f5}
\end{figure*}

@@

To test our hypotheses, we estimated a series of two-parameter logistic
item response theory models with participants' responses to the question
whether they thought each action was an acceptable means of protest as
the outcome variable. Our models differed from the models used in
Experiment 1 in two ways. First, we estimated the item parameters
(\(\alpha_i\), \(\beta_i\)) as \emph{correlated} varying effects.
Second, we used varying, rather than fixed, effects to estimate
differences between the eight conditions. This resulted in partial
pooling---condition-wise estimates were shrunk towards each other---and
allowed us to compare conditions without multiple comparison problems
(Gelman et al., 2012). Our models assigned weakly informative prior
distributions to all model parameters.\footnote{Our model assigned
  \(\text{LKJ} (2)\) prior distributions to the Cholesky-transformed
  correlation matrices for varying effects and
  \((\text{Half-})\text{Cauchy} (0, 5)\) prior distributions to all
  other model parameters.} Figure \ref{fig:f4} shows results from three
preregistered models we estimated to test our hypotheses.

@@

Model 3 extended Model 1 by estimating participants' responses as a
function of their self-reported support for the cause of the protest. We
recoded participants' responses to create a predictor variable that
encoded support for defunding the police when protesters supported
defunding the police and opposition to defunding the police when
protesters opposed defunding the police. As preregistered, we modeled
this relationship as a monotonic effect (Bürkner \& Charpentier, 2020)
that estimated the average change in the outcome variable across
predictor categories as well as how much of this change occurred between
each of the four pairs of adjacent predictor categories.\footnote{Our
  model assigned a Dirichlet prior, \(\alpha = {1, 1, 1, 1}\), to the
  proportions of the overall change that was expected to occur between
  each of the four pairs of predictor categories.} Participants who
supported the protesters' cause did not, on average, consider the same
protest actions to be more acceptable than participants who opposed the
protesters' cause (\(\beta_{y} = 0.14, [-0.12, 0.40]\)). Ergo, our
results did not support the alternative hypothesis that ideology-based
double standards can be reduced to support for, or opposition to, the
cause of a protest.

@@

We included additional measures, not used in the analyses reported in
this paper, including measures of moral conviction (Ryan, 2014),
gender-related system-justifying beliefs (Jost \& Kay, 2005), and the
updated moral foundations questionnaire (Atari et al., 2022, for
details, see SOM). Within the latter questionnaire, we embedded three
further attention checks (e.g., ``To show that you are paying attention
and giving your best effort, please select `moderately describes
me'.'').

@@

\begin{figure*}[!t]
\centering
\caption{Results from the preregistered analyses for Experiment 3}
\includegraphics[scale=1]{../Experiment 3/figures/figure-6}
\caption*{\textit{Note.} Results show the estimated \textit{z}-standardized tendency ($\theta_{(z)}$) to consider more controversial actions acceptable means to protest for or against restricting abortion as a function of the participants' ideology and gender.}
\label{fig:f6}
\end{figure*}

@@

Contradicting our hypothesis of group-based double standards,
participants in three preregistered experiments did not show consistent
ingroup bias in terms of social class when judging protests for workers'
rights (Experiment 1), in terms of race when judging protests for and
against defunding the police (Experiment 2), and in terms of gender when
judging protests for and against restricting abortion (Experiment 3).
Instead, we found that members of advantaged groups---middle-class
(Experiment 1), White (Experiment 2), and male (Experiment 3)
participants---tended to consider all protest actions more acceptable
than members of relatively disadvantaged groups. That is, our findings
showed that participants' judgements of protest actions depended on
their group memberships---but not in the directions implied by the
hypothesized group-based double standards.

@@

In the remainder of this General Discussion, we discuss limiting
conditions on our findings as well as theoretical and practical
implications for understanding the often divided response to collective
action.

\hypertarget{limitations}{%
\subsection{Limitations}\label{limitations}}

Despite its strengths, our research has several limitations that
constrain the generalizability of our findings. First, our research
examined the hypothesized relationships for three political
causes---protecting workers' rights, defunding the police, restricting
abortion---and thus provided only limited evidence that our findings
generalize beyond those causes. Relying on few stimuli to establish an
effect threatens the validity and replicability of research findings and
constrains the generalizability of psychological research (Yarkoni,
2022). Future research should sample a wider range of causes to address
this pervasive but often ignored problem (Judd et al., 2012). This is
particularly important when studying collective action as recent
theorizing (Jost et al., 2017) and research (Osborne et al., 2019)
highlighted the importance of differentiating between progressive and
conservative causes. Second, our research was based on samples from two
Western, Educated, Industrialized, Rich, and Democratic (WEIRD, Henrich
et al., 2010) contexts---the United Kingdom and the United
States---which limits the cross-cultural generalizability of our
findings.

>>>>

\noindent Collective action is a powerful force driving social change
but often sparks contention about what actions are acceptable means to
effect social change. We investigated double standards in judging
collective action---that is, whether observers judge the same protest
actions to be more acceptable depending on who the protesters are and
what they are protesting. In two studies, we used item response theory
to develop an instrument of 25 controversial protest actions to measure
where people draw the line between acceptable and unacceptable forms of
collective action. In three preregistered experiments (\(N = 2776\)), we
found no consistent evidence for ingroup bias in terms of social class
when judging protests for workers' rights (Experiment 1), in terms of
race when judging protests for and against defunding the police
(Experiment 2), and in terms of gender when judging protests for and
against restricting abortion (Experiment 3). Instead, we found that
progressive participants (Experiments 1--3) who rejected
system-justifying beliefs (Experiments 1--2) considered the same protest
actions more acceptable when a cause aligned with their ideological
orientation (for workers' rights, for defunding the police, against
restricting abortion) than when it did not (against defunding the
police, for restricting abortion). Conservative participants considered
the same actions somewhat more acceptable when protesters supported,
rather than opposed, restricting abortion (Experiment 3) but considered
all protest actions, for and against defunding the police, equally
unacceptable (Experiment 2). Our findings have theoretical and practical
implications for understanding the often-divided response to social
movements.

@@

\begin{center}
\noindent 12,257 words
\end{center}

@@

\noindent This research shows that what people consider an acceptable
\emph{means} of protest depends on what the \emph{end} of a protest is.
That is, observers sometimes judge the \emph{same} controversial protest
action (e.g., blocking a road) to be more acceptable when the
protesters' cause (e.g., defunding the police) aligns with their own
ideological position. Conservatives, who support social order and oppose
social change, tend to consider all protest actions less acceptable than
progressives. Progressives, in turn, tend to consider disruptive but
non-violent actions acceptable means to achieve social change and
challenge social injustice. These findings help explain the
often-divided response to recent protest movements.

@@

Another explanation for ideology-based double standards is that they are
rooted in ideological differences in how conservatives and progressives
relate to social change and social inequality which, in turn, are
motivated by basic psychological needs (Jost et al., 2003; Jost, 2021).
As motivated social cognition, ideological differences shape how we
think and feel about political issues, actions, and events. In this
view, ideology-based double standards depend on whether the cause of a
protest aligns with someone's ideological positions.

@@

Item response theory is a conceptual and statistical framework for
understanding how the characteristics of both items and respondents
shape responses to a set of items (DeMars, 2010). Applied to judging
protest actions, this framework allows us to disentangle how
\emph{acceptable} different protest actions are from how
\emph{accepting} different respondents are of them. In line with the
long-standing distinction between normative and non-normative collective
action (Wright et al., 1990), it recognizes that some protest actions
(e.g., signing petitions) are more acceptable than others (e.g.,
blocking roads). In line with the hypothesized double standards, it
recognizes that some people consider more protest actions acceptable
than others. By comparing how accepting respondents are of the
\emph{same} protest actions in \emph{different} situations, we can use
item response theory to detect double standards in judging collective
action.

In this research, respondents judge whether each of a set of more and
less controversial protest actions is an acceptable means of protest (1
= \emph{yes}, 0 = \emph{no}). We use two-parameter item response theory
models to estimate the probability \(\Pr ( y_\textit{ij} = 1 )\) that
respondent \(j\) considers action \(i\) an acceptable means of protest
as a function of three latent model parameters:
\[ \Pr ( y_\textit{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) \]
where \(\theta_j\) estimates how \emph{accepting} respondent \(j\) is of
various actions, \(\beta_i\) estimates how \emph{acceptable} action
\(i\) is, and \(\alpha_i\) estimates how \emph{discriminating} action
\(i\) is.

\begin{figure}
\centering
\caption{Item response curves for three hypothetical protest actions}
\includegraphics[scale=1]{../Scale Development/figures/figure-1.png}
\caption*{\textit{Note.} $\Pr ( y_\textit{ij} = 1 )$ is the probability that that respondent $j$ considers action $i$ an acceptable means of protest. $\theta_{(z)}$ is the $z$-standardized tendency for a participant to be more accepting of various actions. It follows a standard normal distribution and is shown as both $z$-scores (bottom) and percentiles (top). $\beta_i$ and $\alpha_i$ capture, respectively, how acceptable and discriminating an action is.}
\label{fig:f1}
\end{figure}

Figure \ref{fig:f1} illustrates those relationships for three
hypothetical protest actions. \(\theta_{(z)}\) is the
\emph{z}-standardized tendency for a participant to be more accepting of
various actions. It follows a normal distribution so that
\(\theta_{(z)} = 1.28\) means that a respondent is more accepting than
90\% of respondents. \(\beta_i\) determines the probability that the
average respondent (\(\theta_{(z)} = 0\)) considers an action an
acceptable means of protest. Action 2 (\(\beta_2 = 1.28\)) is more
acceptable than Actions 1 and 3 (\(\{ \beta_1, \beta_3 \} = 0\)) as the
average respondent is more likely to consider the former (\(\Pr = .78\))
than the latter (\(\Pr = .50\)) acceptable means of protest.
\(\alpha_i\) determines how much the probability that a respondent
considers an action acceptable depends on how accepting that respondent
is. Action 3 (\(\alpha_3 = 3.60\)) is more discriminating than Actions 1
and 2 (\(\{ \alpha_1, \alpha_2 \} = 1\)) as the slope for the
relationship between how accepting a respondent is and how likely they
are to consider an action acceptable is steeper for the former action.
This means that knowing whether someone considers Action 3 acceptable
provides more information about them than knowing whether they consider
Actions 1 or 2 acceptable.

In this research, item response theory serves two purposes. First, it
provides a statistical framework for scale development. In Studies 1 and
2, we select the most discriminating and informative actions to build an
instrument for capturing double standards in judging collective action.
Second, it provides a formal and statistical definition of double
standards in judging collective action as respondents being more
accepting of the same protest actions depending on who the protesters
are and what they are protesting. In Experiments 1--3, we operationalize
double standards as differences in \(\theta_{(z)}\) which, as
\(\theta_{(z)}\) is \(z\)-standardized, correspond to Cohen's \(d\)
effect sizes. As \(\theta_{(z)}\) determines the probability of
respondents judging different protest actions acceptable, it can be
converted to the predicted proportion \(p\) and number \(n\) of actions
a respondent considers acceptable. Figure \ref{fig:f3} (A) illustrates
these relationships. Figures \ref{fig:f3} (B), \ref{fig:f4},
\ref{fig:f5}, and \ref{fig:f6} show the predicted proportion of actions
a participant would consider acceptable as an intuitive and easily
interpretable effect size.

@@

\noindent Participants were then asked to name between five and ten
actions that fit that definition. Participants were encouraged to think
of actions that varied in how acceptable they were in their opinion. We
recoded responses into a smaller set of unique actions, which we
supplemented with protest actions from the psychological and political
science literature (e.g., Sharp, 1973). This process resulted in 72
actions that varied in how acceptable we would expect them to be. For
details, see Supplemental Online Material A (SOM-A).

@@

We estimated a graded response model (Bürkner, 2021; Samejima,
1997)---an item response theory model for ordinal response
variables---for participants' ratings of how often an action would be an
acceptable form of collective action.\footnote{Ratings of how often a
  given action would be an acceptable form of collective action were
  strongly and negatively correlated with how disruptive, violent, and
  extreme participants considered the action to be and how negative they
  felt about it (Table S3).} Based on the results, we selected the 25
most discriminating, informative, and relevant protest actions to form
an instrument for measuring double standards in judging collective
action in Experiments 1--3. We selected controversial, and thus
diagnostic, actions (e.g., disrupting traffic) over actions that almost
all respondents considered acceptable (e.g., signing petitions) or
unacceptable (e.g., political violence; Figure \ref{fig:f2}). For
detailed ratings and results, see SOM-A and Tables S1--S3.

@@

By varying the participants' and the protesters' group memberships, we
tested the preregistered hypothesis that participants would judge the
same protest actions as more acceptable if the protesters were ingroup
members protesting for the ingroup's rights than if they were outgroup
members protesting for the outgroup's rights. By comparing reactions to
working-class and middle-class protesters, we further tested how
observers judge collective action by lower-status (i.e.~working-class)
and higher-status (middle-class) group members. In non-preregistered
analyses, we tested whether participants would judge the same protest
actions as more or less acceptable depending on their own ideological
positions in terms of political orientation, system-justifying beliefs,
and social dominance orientation.\footnote{In Experiment 1, we explored
  a broader range of operationalizations of political ideology,
  including social dominance orientation, than in the other experiments.}
In this way, Experiment 1 provided a first test of the hypothesized
class- and ideology-based double standards in judging collective action.

@@

We preregistered the sample size as well as all hypotheses,
inclusion/exclusion criteria, measures, and manipulations
(\url{https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b}).
For Experiments 1 and 2, we had preregistered the additional hypothesis
that all participants would judge the same actions to be more acceptable
when the protesters were from the higher-status group rather than the
lower-status group (Experiment 1) and when the goal of the protest was
to support rather than challenge the system (Experiment 2). This
hypothesis reflected the assumption that the motivation to justify the
system was, to some extent, universal. With hindsight, we no longer
think that this assumption is tenable and, therefore, do not discuss the
hypothesis based on it. We still follow the same preregistered analysis
plan and report all results related to our original hypotheses.

We made all materials, data, and analysis scripts available online
(\url{https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62}).
We followed sample size recommendations for item response theory models
(DeMars, 2010, p. 36), planning to recruit 500 participants.

@@

In the screening survey, participants read an accessible definition of
what working-class and middle-class/professional jobs are (for details,
see SOM-B). Participants then answered, among other questions, whether
they considered their current job---or the jobs they had had in the past
or expected to have in the future---to be a working-class job or a
middle-class/professional job. As preregistered, we excluded
participants from the lower-status group who did not respond
``working-class job'' and participants from the higher-status group who
did not respond ``middle-class/professional job''.

@@

\begin{figure*}[!t]
\centering
\caption{Results from the preregistered (\textbf{A}) and non-preregistered (\textbf{B}) analyses for Experiment 1}
\includegraphics[scale=1]{../Experiment 1/figures/figure-3}
\caption*{\textit{Note.} HS = Higher Status, LS = Lower Status. (\textbf{A}) $\theta_{(z)}$ is the \textit{z}-standardized tendency to consider more controversial actions acceptable means of protest. $\Pr$ and $n$ are, respectively, the predicted proportion and number of actions a participant would consider acceptable means of protest in each condition. Bars enclose the 95\% most plausible estimates. (\textbf{B}) Ribbons enclose, from the darkest to the lightest shade, the 50\%, 80\%, 95\%, and 99\% most plausible estimates.}
\label{fig:f3}
\end{figure*}

@@

We estimated this model in CmdStan (Gabry \& Cesnovar, 2021; Stan
Development Team, 2021) using Bayesian statistical methods. Bayesian
inference involves choosing a likelihood function and prior
distributions. A likelihood function links the observed data to the
model parameters and states how likely the observed data are given
different values of said model parameters. Prior distributions state how
plausible different values of said model parameters are before
considering the observed data. Bayesian inference applies Bayes' theorem
to update prior distributions in light of the observed data to produce
posterior distributions. Unlike \(p\)-values and confidence intervals,
posterior distributions have a straight-forward interpretation as
stating how plausible different values of the model parameters are given
the observed data. We report point estimates, based on the median of
posterior samples, and uncertainty intervals, based on the quantiles of
posterior samples, that enclose the 95\% most plausible estimates.

Our model derived the likelihood of participants' responses from a
Bernoulli likelihood function with a logistic regression equation
linking the two item parameters (\(\alpha_i\), \(\beta_i\)), the one
participant parameter (\(\theta_j\)), and the three regression
coefficients to the observed data. To identify the model, we constrained
\(\theta_j\) to have a mean of zero and constrained \(\alpha_i\) to have
a fixed mean and to be non-negative. We used partial pooling to estimate
\(\alpha_i\), \(\beta_i\), and \(\theta_j\). Our model assigned weakly
informative prior distributions (Gelman et al., 2017) to all model
parameters: \(\text{Half-Cauchy} (0, 3)\) prior distributions to the
standard deviations of \(\alpha_i\), \(\beta_i\), and \(\theta_j\) and
\(\text{Normal} (0, 3)\) prior distributions to all other model
parameters.

We report differences between conditions in terms of differences in the
estimated tendency for a participant to consider more protest actions
acceptable (\(\theta_{(z)}\)) which, as \(\theta_{(z)}\) is
\(z\)-standardized, correspond to Cohen's \(d\) effect sizes. Likewise,
we report standardized regression coefficients (\(\beta_{xy}\)) that
correspond to the estimated change in the \(z\)-standardized tendency to
consider more protest actions acceptable for each \(+1\textit{SD}\)
change in the predictor variable. All figures show the predicted
proportion of protest actions participants would consider acceptable in
each condition based on the estimated models.

@@

Figure \ref{fig:f3} shows (B) how participants' tendency to consider
more controversial protest actions acceptable varied as a function of
the three ideological orientation variables. We found that participants
who reported a more right-wing political orientation
(\(\beta_{xy} = -0.25, [-0.37, -0.15]\)) and who expressed more
agreement with system-justifying beliefs
(\(\beta_{xy} = -0.20, [-0.30, -0.10]\)) tended to find fewer collective
actions to be acceptable. In contrast, we found that, after controlling
for the other two variables, social dominance orientation was not
associated with participants' judgments about how acceptable various
collective actions are (\(\beta_{xy} = -0.05, [-0.15, 0.05]\)). Overall,
these findings suggest that people who are right-wing and endorse
system-justifying beliefs tend to find various collective actions to be
less acceptable than people who are left-wing and do not endorse
system-justifying beliefs.\footnote{We ran additional analyses to
  explore how these associations differed across experimental
  conditions. We found uncertainty intervals for these associations to
  overlap across conditions, suggesting that we did not have enough data
  to differentiate these varying effects.}

@@

\begin{figure*}[!t]
\centering
\caption{Results from the preregistered analyses for Experiment 2}
\includegraphics[scale=1]{../Experiment 2/figures/figure-4}
\caption*{\textit{Note.} Against = Protesters oppose defunding the police. For = Protesters support defunding the police. $\Pr$ is the predicted proportion of actions a participant would consider acceptable means of protest. Ribbons enclose, from the darkest to the lightest shade, the 50\%, 80\%, 95\%, and 99\% most plausible estimates.}
\label{fig:f4}
\end{figure*}

\begin{figure*}[!t]
\centering
\caption{Predictions from the preregistered (system justification) and non-preregistered (political orientation) analyses for Experiment 2}
\includegraphics[scale=1]{../Experiment 2/figures/figure-5}
\caption*{\textit{Note.} Against = Protesters oppose defunding the police. For = Protesters support defunding the police. $\Pr$ is the predicted proportion of actions a participant would consider acceptable means of protest in each condition if they were higher ($+1\text{SD}$) or lower ($-1\text{SD}$) than average in system justification (left) or political conservatism (right).}
\label{fig:f5}
\end{figure*}

@@

To test our hypotheses, we estimated a series of two-parameter logistic
item response theory models with participants' responses to the question
whether they thought each action was an acceptable means of protest as
the outcome variable. Our models differed from the models used in
Experiment 1 in two ways. First, we estimated the item parameters
(\(\alpha_i\), \(\beta_i\)) as \emph{correlated} varying effects.
Second, we used varying, rather than fixed, effects to estimate
differences between the eight conditions. This resulted in partial
pooling---condition-wise estimates were shrunk towards each other---and
allowed us to compare conditions without multiple comparison problems
(Gelman et al., 2012). Our models assigned weakly informative prior
distributions: \(\text{LKJ} (2)\) prior distributions to the
Cholesky-transformed correlation matrices for varying effects and
\((\text{Half-})\text{Cauchy} (0, 5)\) prior distributions to all other
model parameters. Figure \ref{fig:f4} shows results from three
preregistered models we estimated to test our hypotheses.

@@

Model 3 extended Model 1 by estimating participants' responses as a
function of their self-reported support for the cause of the protest. We
recoded participants' responses to create a predictor variable that
encoded support for defunding the police when protesters supported
defunding the police and opposition to defunding the police when
protesters opposed defunding the police. As preregistered, we modeled
this relationship as a monotonic effect (Bürkner \& Charpentier, 2020)
that estimated the average change in the outcome variable across
predictor categories as well as how much of this change occurred between
each of the four pairs of adjacent predictor categories. This model
assigned a Dirichlet prior, \(\alpha = {1, 1, 1, 1}\), to the
proportions of the overall change that was expected to occur between
each of the four pairs of predictor categories. We found that
participants who supported the protesters' cause did not, on average,
consider the same protest actions to be more acceptable than
participants who opposed the protesters' cause
(\(\beta_{y} = 0.14, [-0.12, 0.40]\)). Ergo, our results did not support
the alternative hypothesis that ideology-based double standards can be
reduced to support for, or opposition to, the cause of a protest.

@@

We included additional measures, not used in the analyses reported in
this paper, including measures of moral conviction (Ryan, 2014),
gender-related system-justifying beliefs (Jost \& Kay, 2005), and the
updated moral foundations questionnaire (Atari et al., 2023, for
details, see SOM-C). Within the latter questionnaire, we embedded three
further attention checks (e.g., ``To show that you are paying attention
and giving your best effort, please select `moderately describes
me'.'').

@@

\begin{figure*}[!t]
\centering
\caption{Results from the preregistered analyses for Experiment 3}
\includegraphics[scale=1]{../Experiment 3/figures/figure-6}
\caption*{\textit{Note.} Results show the predicted proportion ($\Pr$) of actions a participant would consider acceptable means of protest for or against restricting abortion as a function of the participants' ideology and gender.}
\label{fig:f6}
\end{figure*}

@@

Contradicting our hypothesis of group-based double standards,
participants in three preregistered experiments did not show consistent
ingroup bias in terms of social class when judging protests for workers'
rights (Experiment 1), in terms of race when judging protests for and
against defunding the police (Experiment 2), and in terms of gender when
judging protests for and against restricting abortion (Experiment 3).
Instead, we found that members of advantaged groups---middle-class
(Experiment 1), White (Experiment 2), and male (Experiment 3)
participants---tended to consider all protest actions more acceptable
than members of relatively disadvantaged groups. That is, our findings
showed that participants' judgments of protest actions depended on their
group memberships---but not in the directions implied by the
hypothesized group-based double standards.

@@

In the remainder of this General Discussion, we discuss theoretical and
practical implications for understanding the often divided response to
collective action as well as limitations that constraint the generality
of our findings.
